---
title: "Stroke Data Analysis"
author: "Kai Wilson"
date: "2024-10-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(cluster)    # For clustering algorithms
library(factoextra) # For clustering visualization
```

## Stroke Data Analysis 

In this project, I obtained this dataset from [Kaggle](https://www.kaggle.com/datasets/fedesoriano/stroke-prediction-dataset?resource=download).The data is based on people that have and haven't had a stroke. The dataset was used for training a model to make predictions if a person is likely to have a stoke. 

The data is based on people that have and haven't had a stroke. The dataset was used for training a model to make predictions if a person is likely to have a stroke.


After I collected the data, cleaned it, and put it into a SQLite database. For the database schema, I made one table for person and another table for their medical history.


I queried the data through joining the two tables and completing my analysis. After I did three queries, I turned them into their own CSV files to quickly graph the data using ggplot. Taking a step further, I created two K-means clusters: one focusing on just the medical history and another with medical history based on gender.


All code can be found on my [github](https://github.com/Kwilso3412/medical_data_analysis): https://github.com/Kwilso3412/medical_data_analysis




## Totals





Had a Stroke: 249 


Havent had a Stroke: 4861\
 
 
 
 
Men that have had a stroke: 108 


Men married had stroke: 100\



 
Women that have had a stroke: 141 

Women married had stroke: 120 


```{r Glucose, echo=FALSE, warning=FALSE, message=FALSE}
glucose = read.csv("glucose.csv")
# Remove N/A values 
glucose = filter(glucose, !grepl("Other", gender))
glucose$age_group = cut(glucose$age, breaks = seq(0, 90, by = 5))  # group by 5-year intervals
avg_by_group = aggregate(avg_glucose_level ~ age_group + gender, data = glucose, FUN = mean)


ggplot(avg_by_group, aes(x = age_group, y = avg_glucose_level, color = gender)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Age vs. Glucose Level by Gender",
       x = "Age",
       y = "Average Glucose Level") +
  scale_color_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),  # Adjust text angle and position
    plot.margin = margin(b = 40, l = 40),  # Add margin at bottom
    aspect.ratio = 0.5  # Make plot wider (decrease for wider plot)
  ) +
  coord_cartesian(expand = TRUE)  # Expand the plot area

```
The graph shows that blood glucose levels tend to increase with age for both men and women. While levels stay fairly stable until age 30, they begin rising after that, with a sharper increase after age 50. Men (blue line) generally show higher glucose levels than women (pink line), with this difference becoming more noticeable in older age groups, particularly after age 60.



```{r BMI, echo=FALSE, warning=FALSE}
bmi = read.csv("bmi.csv")
bmi = filter(bmi, !grepl("N/A", bmi))
bmi = filter(bmi, !grepl("Other", gender))
bmi$age_group = cut(bmi$age, breaks = seq(0, 90, by = 5))  # group by 5-year intervals
avg_by_group = aggregate(bmi ~ age_group + gender, data = bmi, FUN = mean)
bmi$age = round(bmi$age, 1)


ggplot(avg_by_group, aes(x = age_group, y = bmi, color = gender, group = gender)) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(title = "Chance of Stroke by BMI",
       x = "Age Group",
       y = "Average BMI") +
  scale_color_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # rotate x-axis labels
```
This graph shows the relationship between age and BMI (body mass index) for men and women. BMI increases with age until around 45-50 years old, then gradually decreases in later years. Both genders follow similar patterns, with slight variations but no major differences between men and women.



```{r Residence, echo=FALSE, warning=FALSE}
residence = read.csv("residence.csv")
residence = filter(residence, !grepl("Other", gender))
residence$age = round(residence$age, 1)
residence$age_group = cut(residence$age, breaks = seq(0, 90, by = 5))  # group by 5-year intervals


ggplot(residence, aes(x = residence_type, y = age, fill = gender)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  theme_minimal() +
  labs(title = "Average Age by Residence Type and Gender",
       x = "Residence Type",
       y = "Average Age") +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  stat_summary(fun = mean, geom = "text", 
              aes(label = round(..y.., 1)), 
              position = position_dodge(width = 0.9),
              vjust = -0.5)

```
This graph shows the average age of people living in rural versus urban areas, broken down by gender. The differences are small, but urban women have the highest average age (44.2 years), while rural men have the lowest (42.4 years). Women in both locations are slightly older on average than men.


\newpage
## Clustering Based on Medical History



Cluster 1 (Red):

Primarily young individuals (average age ~29.4)
Healthy BMI range (~25.6)
Normal glucose levels (~98.2)
Very low rates of hypertension (2.6%) and heart disease (2.3%)
All marked as unknown smoking status
Represents the "healthy young adult" group

Cluster 2 (Blue):

Oldest group (average age ~51.2)
Highest BMI (30.7, classified as obese)
Highest glucose levels (~109.8)
Highest rates of hypertension (12.4%) and heart disease (8.1%)
Mixed smoking status: 53% former smokers, 46.7% current smokers
Represents the "high-risk" group

Cluster 3 (Green):

Middle-aged group (average age ~46.5)
Elevated BMI (30.0, borderline obese)
Elevated glucose levels (~107.1)
Moderate hypertension (11.7%) and heart disease (4.4%)
All never-smokers
Represents the "moderate-risk" group\

```{r Kmeans Cluster numeric, echo=FALSE, warning=FALSE}
full_data = read.csv("healthcare_dataset_stroke_data.csv", na.strings = c("", "NA", "N/A"))
full_data = filter(full_data, !grepl("Other", gender))
# Create dummy variables using model.matrix
smoking_dummies = model.matrix(~ smoking_status - 1, data = full_data)
# Convert to data frame
smoking_df = as.data.frame(smoking_dummies)
# Clean up column names
names(smoking_df) = gsub("smoking_status", "", names(smoking_df))
names(smoking_df) = gsub("`", "", names(smoking_df))
names(smoking_df) = gsub(" ", "_", names(smoking_df))

# Remove original smoking_status column and add dummy columns
full_data = cbind(full_data[, !names(full_data) %in% "smoking_status"], smoking_df)

full_data$age = round(full_data$age, 1)
full_data$age_group = cut(full_data$age, breaks = seq(0, 90, by = 5))

numeric_clustering = function(full_data) {
    numeric_data = full_data[, c("age", "avg_glucose_level", "bmi","hypertension","heart_disease","formerly_smoked","never_smoked","smokes","Unknown")]
    complete_data = na.omit(numeric_data)
    scaled_data = scale(complete_data)
    wss = numeric(10)
    for(i in 1:10) {
        km <- kmeans(scaled_data, centers = i, nstart = 25)
        wss[i] = km$tot.withinss
    }
    
    k = 3 
    km_result = kmeans(scaled_data, centers = k, nstart = 25)
    result_data = full_data
    result_data$numeric_cluster = NA
    complete_rows = complete.cases(numeric_data)
    result_data$numeric_cluster[complete_rows] <- km_result$cluster
    
    return(list(
        clusters = km_result,
        data = result_data,
        n_excluded = sum(!complete_rows)
    ))
}

results = numeric_clustering(full_data)
numeric_clustering = function(full_data) {
    numeric_data = full_data[, c("age", "avg_glucose_level", "bmi","hypertension","heart_disease","formerly_smoked","never_smoked","smokes","Unknown")]
    complete_data = na.omit(numeric_data)
    scaled_data <- scale(complete_data)
    wss = numeric(10)
    for(i in 1:10) {
        km = kmeans(scaled_data, centers = i, nstart = 25)
        wss[i] = km$tot.withinss
    }
    k <- 3 
    km_result = kmeans(scaled_data, centers = k, nstart = 25)
    result_data = full_data
    result_data$numeric_cluster = NA
    complete_rows = complete.cases(numeric_data)
    result_data$numeric_cluster[complete_rows] = km_result$cluster
    
    return(list(
        clusters = km_result,
        data = result_data,
        n_excluded = sum(!complete_rows)
    ))
}
results = numeric_clustering(full_data)
cluster_summary = aggregate(
    cbind(
        count = 1,
        age = results$data$age,
        avg_glucose_level = results$data$avg_glucose_level,
        bmi = results$data$bmi,
        hypertension = results$data$hypertension,
        heart_disease = results$data$heart_disease,
        formerly_smoked = results$data$formerly_smoked,
        never_smoked = results$data$never_smoked,
        smokes = results$data$smokes,
        Unknown = results$data$Unknown
    ),
    by = list(cluster = results$data$numeric_cluster),
    FUN = function(x) if(length(x) == 1) length(x) else mean(x, na.rm = TRUE)
)
print("Cluster Summary Statistics:")
print(cluster_summary)

```
\newpage
```{r Kmeans Cluster Age vs BMI, echo=FALSE, warning=FALSE}
p1 = ggplot(results$data, aes(x=age, y=bmi, color=factor(numeric_cluster))) +
    geom_point(alpha=0.6) +
    theme_minimal() +
    labs(title="Clustering Results: Age vs BMI",
         x="Age",
         y="BMI",
         color="Cluster") +
    scale_color_brewer(palette="Set1")
print(p1)

```
\


Age vs BMI:


Shows clear age-related BMI increases
Younger cluster (1) maintains healthier BMI range
More variation in BMI as age increases


```{r Kmeans Cluster glucose vs BMI, echo=FALSE, warning=FALSE}
p2 = ggplot(results$data, aes(x=bmi, y=avg_glucose_level, color=factor(numeric_cluster))) +
    geom_point(alpha=0.6) +
    theme_minimal() +
    labs(title="Clustering Results: BMI vs Glucose Level",
         x="BMI",
         y="Average Glucose Level",
         color="Cluster") +
    scale_color_brewer(palette="Set1")
print(p2)

```
\


BMI vs Glucose Level:


Shows positive correlation between BMI and glucose levels
Higher BMI associated with more variable glucose levels
Clear separation between healthy (Cluster 1) and at-risk groups (Clusters 2 and 3)


```{r Kmeans Cluster Age vs glucose, echo=FALSE, warning=FALSE}

p3 = ggplot(results$data, aes(x=age, y=avg_glucose_level, color=factor(numeric_cluster))) +
    geom_point(alpha=0.6) +
    theme_minimal() +
    labs(title="Clustering Results: Age vs Glucose Level",
         x="Age",
         y="Average Glucose Level",
         color="Cluster") +
    scale_color_brewer(palette="Set1")
print(p3)

```
\


Age vs Glucose Level:


Strong positive correlation between age and glucose levels
More scattered/variable glucose levels in older age groups
Cluster 1 (young) shows tighter grouping of normal glucose levels



### In Summary

Age is a strong determining factor in health outcomes
Risk factors tend to cluster together (high BMI, glucose, hypertension, heart disease)
Smoking status appears to be an important discriminating factor between clusters
Clear progression from healthy young adults to higher-risk older adults
Potential for early intervention in younger populations to prevent progression to higher-risk clusters
\newpage

## Clustering Based on Gender: Gender Distribution in Clusters

Cluster 1:

More balanced gender distribution
Approximately 750 females and 600 males
This was the "younger, healthier" cluster with lower BMI, lower glucose levels, and minimal health risks

Cluster 2:

Predominantly male
Approximately 1400 males
Very few or no females
This represented the "higher risk" group with higher BMI, glucose levels, and cardiovascular risks

Cluster 3:

Predominantly female
Approximately 2000 females
Very few or no males
This represented the female equivalent of Cluster 2, with similar age and health risk patterns\


```{r Kmeans Cluster by gender distribution, echo=FALSE, warning=FALSE}
gender_clustering = function(full_data) {
    gender_dummy = model.matrix(~ gender - 1, data = full_data)
    numeric_data = full_data[, c("age", "avg_glucose_level", "bmi",
                                 "hypertension", "heart_disease",
                                 "formerly_smoked", "never_smoked",
                                 "smokes", "Unknown")]
    combined_data = cbind(numeric_data, gender_dummy)
    complete_data = na.omit(combined_data)
    scaled_data = scale(complete_data)
    wss = numeric(10)
    for(i in 1:10) {
        km = kmeans(scaled_data, centers = i, nstart = 25)
        wss[i] = km$tot.withinss
    }
    k = 3  
    km_result = kmeans(scaled_data, centers = k, nstart = 25)
    result_data = full_data
    result_data$numeric_cluster = NA
    complete_rows = complete.cases(combined_data)
    result_data$numeric_cluster[complete_rows] = km_result$cluster
    return(list(
        clusters = km_result,
        data = result_data,
        n_excluded = sum(!complete_rows)
    ))
}

results = gender_clustering(full_data)
filtered_data = subset(results$data, !is.na(numeric_cluster))
cluster_summary = aggregate(
    data.frame(count = rep(1, nrow(filtered_data))),
    by = list(
        numeric_cluster = filtered_data$numeric_cluster,
        gender = filtered_data$gender
    ),
    FUN = sum
)

gender_dist = ggplot(cluster_summary, 
                     aes(x=factor(numeric_cluster),
                         y=count, 
                         fill=gender)) +
    geom_bar(stat="identity", position="dodge") +
    theme_minimal() +
    labs(title="Gender Distribution by Cluster",
         x="Cluster",
         y="Count",
         fill="Gender") +
    scale_fill_brewer(palette="Set2")
print(gender_dist)

```
This visualization reveals that the clustering algorithm effectively separated the middle-aged/higher risk population by gender (Clusters 2 and 3), while keeping younger, healthier individuals of both genders together in Cluster 1. This gender segregation in the higher-risk clusters might be useful for developing targeted health interventions specific to each gender's risk patterns.
\newpage

## Clustering Based on Gender

Cluster 1:

Predominantly younger individuals (Female avg: 29.4 years, Male avg: 22.2 years)
Lower BMI (Female: 25.3, Male: 23.9) - both in the normal/slightly overweight range
Lower glucose levels (Female: 94.6, Male: 93.5) - both within normal range
Very low rates of hypertension and heart disease (<1%)
All marked as "Unknown" for smoking status (unknown_mean = 1)
Represents the healthiest cluster overall

Cluster 2/3 (higher risk group):
Females (Cluster 3):

Middle-aged (avg: 48.4 years)
Higher BMI (30.4) - in the obese range
Elevated glucose (106.5)
Higher rates of hypertension (11.6%) and heart disease (4.6%)
Mixed smoking status: 21.4% formerly smoked, 56.3% never smoked, 19.9% current smokers

Males (Cluster 2):

Slightly older (avg: 50.5 years)
Similar BMI (30.7) - also in obese range
Highest glucose levels (114.3)
Highest rates of hypertension (14.1%) and heart disease (10%)
Different smoking pattern: 26.8% formerly smoked, 45.9% never smoked, 22.1% current smokers\

```{r Kmeans Cluster by gender, echo=FALSE, warning=FALSE}
clustered_data <- results$data
cluster_summary <- aggregate(
    cbind(
        n = 1,
        age = clustered_data$age,
        avg_glucose_level = clustered_data$avg_glucose_level,
        bmi = clustered_data$bmi,
        hypertension = clustered_data$hypertension,
        heart_disease = clustered_data$heart_disease,
        formerly_smoked = clustered_data$formerly_smoked,
        never_smoked = clustered_data$never_smoked,
        smokes = clustered_data$smokes,
        Unknown = clustered_data$Unknown
    ),
    by = list(
        cluster = clustered_data$numeric_cluster,
        gender = clustered_data$gender
    ),
    FUN = function(x) if(length(x) == 1) length(x) else mean(x, na.rm = TRUE)
)

names(cluster_summary) <- c("cluster", "gender", "count", 
                          "age_mean", "glucose_mean", "bmi_mean",
                          "hypertension_mean", "heart_disease_mean",
                          "formerly_smoked_mean", "never_smoked_mean",
                          "smokes_mean", "unknown_mean")

cluster_summary <- cluster_summary[!is.na(cluster_summary$cluster), ]

print("Cluster Summary Statistics by Gender:")
print(cluster_summary)
```



```{r Kmeans Cluster by gender age bmi, echo=FALSE, warning=FALSE}

p1 <- ggplot(results$data, aes(x=age, y=bmi, color=factor(numeric_cluster), shape=gender)) +
    geom_point(alpha=0.6) +
    theme_minimal() +
    labs(title="Clustering Results: Age vs BMI by Gender",
         x="Age",
         y="BMI",
         color="Cluster",
         shape="Gender") +
    scale_color_brewer(palette="Set1")
print(p1)

```
This scatter plot shows two main clusters in the Age vs BMI relationship:
Cluster 1 (red) shows younger people (mostly under 40) with lower BMI ranges (mostly between 20-30), suggesting a healthier weight group.
Cluster 2 (green) shows a wider age range with higher BMI values (mostly between 25-40), suggesting an overweight/obese group across different ages.
The pattern suggests that while younger people tend to have lower BMI, there's more variation in BMI as age increases, with some people maintaining healthy weights while others show higher BMI values.



```{r Kmeans Cluster by gender age glucose, echo=FALSE, warning=FALSE}
p2 <- ggplot(results$data, aes(x=age, y=avg_glucose_level, color=factor(numeric_cluster), shape=gender)) +
    geom_point(alpha=0.6) +
    theme_minimal() +
    labs(title="Clustering Results: Age vs Glucose Level by Gender",
         x="Age",
         y="Average Glucose Level",
         color="Cluster",
         shape="Gender") +
    scale_color_brewer(palette="Set1")
print(p2)

```
This scatter plot shows Age vs Glucose Levels clustered into two main groups:
Cluster 1 (red) shows people with normal glucose levels (mostly between 60-120), appearing more commonly in younger ages.
Cluster 2 (green) shows elevated glucose levels (above 120), becoming more common with increasing age, possibly indicating pre-diabetic or diabetic ranges.
The pattern suggests that glucose levels tend to increase with age, and older individuals are more likely to have higher glucose readings. The separation between clusters could help identify potential diabetes risk groups across age ranges.


### In Summary


The data shows clear age-related clustering and notable gender differences in health risks, with males showing higher glucose levels and cardiovascular risks. There are similar BMI patterns across genders within the same age groups. Risk factors tend to cluster together, as higher age correlates with higher BMI and glucose levels. Additionally, smoking status varies significantly by both age group and gender.
